---
title: "CS601C Fall Project One"
output:
  pdf_document: default
  word_document: default
  html_document: default
date: "2024-10-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Peter Treacy


# College

### A
```{r view}

setwd("~/Desktop/CompStatistics")

data <- read.csv("~/Desktop/CompStatistics/college.csv")

```

### B

```{r pressure, echo=FALSE}
head(data)
```

### C
````{r Continued, message=FALSE}
#i
summary(data)
````

````{r continued}
#ii

data <- read.csv("~/Desktop/CompStatistics/college.csv")

num_data <- data[, sapply(data, is.numeric)]

selectnum_data <- num_data[, 1:10]

panel.custom <- function(x, y) {
  points(x, y, pch = 1, col = "black", cex = .1)
  abline(lm(y ~ x), col = "red")  
}
par(mar = c(5, 5, 4, 2))

pairs(selectnum_data,
      panel = panel.custom,             
      main = "Scatterplot Matrix: First 10 Columns"
)
````

````{r iii, message=FALSE}

#iii Private
data$Private <- as.factor(data$Private)

# Boxplot
boxplot(Outstate ~ Private, data = data,
        main = "Outstate Tuition by Private/Public",
        xlab = "Private",
        ylab = "Outstate Tuition",
        col = c("lightblue", "lightgreen"))
````

````{r iv, messages=FALSE}
#iv

# Elite = 'Top10perc' column
data$Elite <- ifelse(data$Top10perc > 50, "Elite", "Non-Elite")

# Elite to a categorical variable
data$Elite <- as.factor(data$Elite)

# Out Total
table(data$Elite)
````

````{r v, message=FALSE}
#v
summary(data$Elite)

boxplot(Outstate ~ Elite, data = data,
        main = "Boxplot of Outstate Tuition by Elite Status",
        xlab = "Elite Status",
        ylab = "Outstate Tuition",
        col = c("lightblue", "lightgreen"))

````

````{r vi, message=FALSE}
#vi
data <- read.csv("~/Desktop/CompStatistics/college.csv", stringsAsFactors = TRUE)

par(mfrow=c(2,2))

hist(data$Outstate, breaks = 10, 
     main = "Outstate Tuition", xlab = "Outstate Tuition", 
     col = "lightblue")

hist(data$Room.Board, breaks = 15, 
     main = "Room and Board Costs", xlab = "Room and Board", 
     col = "darkblue")

hist(data$PhD, breaks = 20, 
     main = "Percentage of PhDs", xlab = "PhD Percentage", 
     col = "lightpink")

hist(data$Top10perc, breaks = 8, 
     main = "Top 10% of High School", xlab = "Top 10% of High School", 
     col = "lightyellow")
````

````{r vii, message=FALSE}
#vii

data <- read.csv("college.csv")

cdata <- data[!is.na(data$Accept) & !is.na(data$Apps) & !is.na(data$Grad.Rate), ]

cdata$Accept_Rate <- (cdata$Accept / cdata$Apps) * 100

model <- lm(Grad.Rate ~ Accept_Rate, data = cdata)

plot(cdata$Accept_Rate, cdata$Grad.Rate, 
     main = "Acceptance Rate vs Graduation Rate",
     xlab = "Acceptance Rate (%)", 
     ylab = "Graduation Rate", 
     pch = 20, col = "black")
abline(model, col = "red")
````

### Conclusion:
You could conclude that colleges with higher acceptance rates tend to have higher graduation rates. 
This might suggest that more inclusive colleges do well at retaining and graduating students.

# Auto

````{r Auto Dataset, message=FALSE}
setwd("~/Desktop/CompStatistics")

auto_data <- read.csv("~/Desktop/CompStatistics/auto.csv")

str(auto_data)
````
### 2.A
Quantitative: mpg, cylinders, displacement, horsepower, weight, acceleration, year.

Qualitative: origin, name.

### 2.B
````{r 2.B, message=FALSE}
range(auto_data$mpg)
range(auto_data$cylinders)
range(auto_data$displacement)
range(auto_data$horsepower)
range(auto_data$weight)
range(auto_data$acceleration)
range(auto_data$year)
````
### 2.C
````{r 2.C, message=FALSE}
auto_data <- read.csv("~/Desktop/CompStatistics/auto.csv")

#mpg
mean(auto_data$mpg)
sd(auto_data$mpg)

#cylinders
mean(auto_data$cylinders)
sd(auto_data$cylinders)

#displacement
mean(auto_data$displacement)
sd(auto_data$displacement)

#horsepower
mean(auto_data$horsepower)
sd(auto_data$horsepower)

#weight
mean(auto_data$weight)
sd(auto_data$weight)

#acceleration
mean(auto_data$acceleration)
sd(auto_data$acceleration)

#year
mean(auto_data$year)
sd(auto_data$year)
````

### 2.D
````{r 2.D, message=FALSE}
auto_data <- read.csv("~/Desktop/CompStatistics/auto.csv")

#range, mean standard deviation for the original dataset
roriginal <- sapply(auto_data[, sapply(auto_data, is.numeric)], range)
moriginal <- sapply(auto_data[, sapply(auto_data, is.numeric)], mean)
sdoriginal <- sapply(auto_data[, sapply(auto_data, is.numeric)], sd)

print(roriginal)
print(moriginal)
print(sdoriginal)
````

````{r 2.inspect, message=FALSE}
#remove 10 rows
rows_to_remove <- sample(1:nrow(auto_data), 10)
auto_data_subset <- auto_data[-rows_to_remove, ]

#range, mean, and standard deviation for the subset
range_subset <- sapply(auto_data_subset[, sapply(auto_data_subset, is.numeric)], range)
mean_subset <- sapply(auto_data_subset[, sapply(auto_data_subset, is.numeric)], mean)
sd_subset <- sapply(auto_data_subset[, sapply(auto_data_subset, is.numeric)], sd)

print(range_subset)
print(mean_subset)
print(sd_subset)
````

````{r 2inspect2, message=FALSE}
#mean (percentage change)
mean_comparison <- 100 * (moriginal - mean_subset) / moriginal
mean_comparison

#standard deviation (percentage change)
sd_comparison <- 100 * (sdoriginal - sd_subset) / sdoriginal
sd_comparison
````

Conclusion
The changes in _original to _subset were fairly minor and did not have a subsational impact on the outputs. This is probably due to how large the original dataset is.

#2.E
````{r 2.E, message=FALSE}
auto_data <- read.csv("~/Desktop/CompStatistics/auto.csv")

library(ggplot2)

lm_mpg_hp <- lm(mpg ~ horsepower, data = auto_data)

# Scatterplot of mpg vs horsepower
plot(auto_data$horsepower, auto_data$mpg,
     main = "MPG vs Horsepower",
     xlab = "Horsepower", ylab = "MPG", pch = 19, col = "darkblue")
abline(lm_mpg_hp, col = "red", lwd = 2)

plot(auto_data$weight, auto_data$mpg,
     main = "MPG vs Weight",
     xlab = "Weight", ylab = "MPG", pch = 19, col = "darkgreen")

plot(auto_data$displacement, auto_data$mpg,
     main = "MPG vs Displacement",
     xlab = "Displacement", ylab = "MPG", pch = 19, col = "lightblue")
````

These plots show negative relationships between mpg and horsepower, weight, and displacement because as these variables increase the mpg tends to decrease. Horsepower is a strong predictor of mpg, as we see with the negative slope in the scatterplot. Displacement's plot shows us that larger engines(cars) typically consume more fuel, leading to lower mpg.

# Boston

### 3.A
````{r Boston Data Set, message=FALSE}
setwd("~/Desktop/CompStatistics")

boston_data <- read.csv("~/Desktop/CompStatistics/boston.csv")
# number of row 
nrow(boston_data) 
# number of columns
ncol(boston_data)
````
Each row represents a single observation, a house or property, in a specific suburb of Boston. Each column in the dataset typically represents a different attribute (feature) of the house or the neighborhood such as crime per capita, or proportion of residential land zoned for lots over 25,000 sq. ft

### 3.B
````{r 3.B, message=FALSE}
boston_data <- read.csv("~/Desktop/CompStatistics/boston.csv")

pairs(boston_data[, c("crim", "lstat", "medv", "tax", "rm")],
      main = "Boston Housing Predictors")
````

### Conclusion
Higher crime rates and lower-status populations are associated with lower home values.

Crime rate (crim) vs. median home value (medv), and possibly crim vs. rm (number of rooms), are likely to show strong negative relationships. Higher crime rates generally correspond to lower home values and smaller homes.

Number of rooms (rm) vs. median home value (medv) typically shows a strong positive correlation, as larger homes are associated with higher home values.

Relationships between property tax (tax) and other variables like crim, rm, and medv may show weak correlations, as taxes can vary independently of home size or crime rates in different areas.

### 3.C
````{r 3.C, message=FALSE}
boston_data <- read.csv("~/Desktop/CompStatistics/boston.csv")

#crime rate vs lower status population
crimlstat <- lm(crim ~ lstat, data = boston_data)

summary(crimlstat)
````

### Conclusion
The relationship between crim and lstat is likely to be significant predictors, indicating that crime rates increase as the percentage of lower-status population increases. The scatterplot and regression model provide a clearer view of this.

### 3.D
````{r 3.D, message=FALSE}
summary(boston_data$crim)
summary(boston_data$tax)

# suburbs with crime rates in the top 10%
high_crime_threshold <- quantile(boston_data$crim, 0.9)  
# 90th percentile
high_crime_suburbs <- boston_data[boston_data$crim > high_crime_threshold, ]

# suburbs with tax rates in the top 10%
high_tax_threshold <- quantile(boston_data$tax, 0.9)  
# 90th percentile
high_tax_suburbs <- boston_data[boston_data$tax > high_tax_threshold, ]

# suburbs with high tax rates
high_tax_suburbs[, c("tax", "medv", "crim")]

hist(boston_data$crim, breaks = 50, main = "Distribution of Crimes",
     xlab = "Crime Rate", col = "lightblue")

hist(boston_data$tax, breaks = 50, main = "Distrubution of Taxes",
     xlab = "Tax Rate", col = "lightgreen")

# scatterplot for Pupil-Teacher Ratio vs Median Home Value
plot(boston_data$ptratio, boston_data$medv,
     main = "Pupil-Teacher Ratio vs Median Home Value",
     xlab = "Pupil-Teacher Ratio (Number of Students per Teacher)", 
     ylab = "Median Home Value (in $1000s)", 
     pch = 19, col = "darkblue", cex = 1.3)
grid()
````

### Conclusion

The range of predictors such as crime rate (crim), tax rate (tax), and pupil-teacher ratio (ptratio) highlight disparities across suburbs. 

Areas with high crime rates are likely less desirable, leading to lower property values and a lower ptratio, while some suburbs maintain very low crime levels, suggesting safer environments.

Tax rates vary widely, reflecting differences in local policies and possibly the quality of public services or schools. Likewise, the range in pupil-teacher ratios suggests disparities in educational resources, where some suburbs benefit from lower ratios and better educational quality, while others face higher ratios, potentially indicating overcrowded schools. 

This variability shows the contrasting living conditions across the Boston area, with some suburbs offering safer environments, better educational opportunities, and potentially higher taxes for improved services.

### 3.E
````{r 3.E, message=FALSE}
#3.E
# number of suburbs along the Charles River

num_suburbs_bound_river <- sum(boston_data$chas == 1)

num_suburbs_bound_river
````
### 3.F
````{r 3.F, message=FALSE}
boston_data <- read.csv("~/Desktop/CompStatistics/boston.csv")

# median pupil-teacher ratio
median_ptratio <- median(boston_data$ptratio)

median_ptratio
````

### 3.G
````{r 3.G, message=FALSE}
# lowest median value of suburban homes
min_medv_row <- which.min(boston_data$medv)

lowest_medv_suburb <- boston_data[min_medv_row, ]
lowest_medv_suburb

````

## Conclusion
The suburb with the lowest median home value (medv) (399) likely suffers from a combination of factors that make it less desirable for homebuyers.

A high crime rate (crim) is the most important negative predictor when homebuyers are viewing a suburban town. The ptratio (pupil-teacher) suggests that the educational oppurtunies in the suburb are limited, which could further qdd to it's negative appeal (children, families, etc)

The lstat (high percentage of lower-status residents) is 30.59, which is far away from the lowest on the Boston Housing list, which is 1.73. This indicates socio-economic challenges, which often correlate with lower demand for housing and lower property values. Fewer rooms per dwelling (rm) 5.453 is the median for this Boston suburb town vs the highest listed at 8.780 which is suggests that the homes are smaller, which is another reason why property values are low. 

This shows how a combination of socio-economic, safety, and educational factors can influence housing markets and 399 would not be viewed as desirable. 

### 3.H
````{r 3.H, message=FALSE}
# number of suburbs with more than 7 rooms
suburb7_rooms <- sum(boston_data$rm > 7)

# the number of suburbs with more than 8 rooms
suburb8_rooms <- sum(boston_data$rm > 8)

# Display the results
suburb7_rooms
suburb8_rooms

````
### Conclusion
Suburbs with an average of more than 8 rooms per dwelling are typically found in wealthier areas. These neighborhoods often feature higher property values, lower crime rates, and access to better school resources, offering a distinct contrast to other regions in the dataset. This analysis sheds light on the characteristics that set these affluent suburbs apart.